---
title: "Project 4 Group 1"
author: "Luka Tumbas"
format: html
editor: visual
---
First of all we will construct our data frame necessary for analysis, with 3 Assets, 3 states and 1000 simulations.
```{r}
# Set seed for reproducibility
set.seed(123)

# Number of simulations
n_simulations <- 1000

# Economic states and probabilities
economic_states <- c("High", "Medium", "Low")
state_probabilities <- c(0.30, 0.50, 0.20)

# Asset returns for each state
returns <- data.frame(
  Asset1 = c(0.08, 0.04, -0.02),
  Asset2 = c(0.12, 0.03, -0.05),
  Asset3 = c(0.05, 0.02, 0.01)
)

# Portfolio weights
weights <- c(0.50, 0.30, 0.20)

# Simulate 1000 economic states using sample function
simulated_states <- sample(economic_states, n_simulations, replace = TRUE, prob = state_probabilities)

# Generate data for all simulations using lapply
simulated_data <- lapply(simulated_states, function(state) {
  state_index <- which(economic_states == state)
  asset_returns <- returns[state_index,]
  portfolio_return <- sum(asset_returns * weights)
  
  # Return as a data frame
  data.frame(State = state, 
             Asset1_Return = as.numeric(asset_returns[1]), 
             Asset2_Return = as.numeric(asset_returns[2]), 
             Asset3_Return = as.numeric(asset_returns[3]), 
             Portfolio_Return = portfolio_return, 
             stringsAsFactors = FALSE)
})

# Convert list of data frames to a single data frame
simulated_df <- do.call(rbind, simulated_data)

# View the first few rows
head(simulated_df)

```
Now that we have our data frame with 1000 states and returns, we can focus on analysis. 

```{r}
# Let us compute the mean return for each Asset and the portfolio as a whole
A1_mean_return <- mean(simulated_df$Asset1_Return)
A2_mean_return <- mean(simulated_df$Asset2_Return)
A3_mean_return <- mean(simulated_df$Asset3_Return)
Portfolio_mean_return <- mean(simulated_df$Portfolio_Return)

#Now let us print this
cat(sprintf("The mean return of Asset1 is %.2f%%\n", A1_mean_return * 100))
cat(sprintf("The mean return of Asset2 is %.2f%%\n", A2_mean_return * 100))
cat(sprintf("The mean return of Asset3 is %.2f%%\n", A3_mean_return * 100))
cat(sprintf("The mean return of the portfolio is %.2f%%\n", Portfolio_mean_return * 100))

#Let's calculate the Asset and portfolio standard deviation now
A1_st_dev <- sd(simulated_df$Asset1_Return)
A2_st_dev <- sd(simulated_df$Asset2_Return)
A3_st_dev <- sd(simulated_df$Asset3_Return)
Portfolio_st_dev <- sd(simulated_df$Portfolio_Return)

#And print this as well,but print some separator in between
cat(rep("*", 39), "\n")
cat(sprintf("Standard deviation of Asset1 returns is %.2f%%\n", A1_st_dev * 100))
cat(sprintf("Standard deviation of Asset2 returns is %.2f%%\n", A2_st_dev * 100))
cat(sprintf("Standard deviation of Asset3 returns is %.2f%%\n", A3_st_dev * 100))
cat(sprintf("Standard deviation of the portfolio returns is %.2f%%\n", Portfolio_st_dev * 100))

```
Now to compute the expected portfolio return
```{r}
A1_expected_return <- sum(returns$Asset1 * state_probabilities)
A2_expected_return <- sum(returns$Asset2 * state_probabilities)
A3_expected_return <- sum(returns$Asset3 * state_probabilities)
Portfolio_expected_return <- A1_expected_return*weights[1]+A2_expected_return*weights[2]+A3_expected_return*weights[3]
#Let us print this as well
cat(sprintf("The expected return of the portfolio is %.2f%%\n", Portfolio_expected_return * 100))
```
When we compare the mean portfolio return (3.75%) and the expected portfolio return(3.77%) we can see that they are basically the same, since we used the same state probabilities to compute them.

Now we can look at the expected standard deviation.
```{r}
var_A1 <- sum((returns$Asset1 - A1_expected_return)^2 * state_probabilities)
var_A2 <- sum((returns$Asset2 - A2_expected_return)^2 * state_probabilities)
var_A3 <- sum((returns$Asset3 - A3_expected_return)^2 * state_probabilities)

# Calculate covariance between assets
cov_A1_A2 <- sum((returns$Asset1 - A1_expected_return) * (returns$Asset2 - A2_expected_return) * state_probabilities)
cov_A1_A3 <- sum((returns$Asset1 - A1_expected_return) * (returns$Asset3 - A3_expected_return) * state_probabilities)
cov_A2_A3 <- sum((returns$Asset2 - A2_expected_return) * (returns$Asset3 - A3_expected_return) * state_probabilities)

# Portfolio variance formula
expected_portfolio_variance <- (weights[1]^2 * var_A1) + (weights[2]^2 * var_A2) + (weights[3]^2 * var_A3) +
  2 * (weights[1] * weights[2] * cov_A1_A2 + weights[1] * weights[3] * cov_A1_A3 + weights[2] * weights[3] * cov_A2_A3)
expected_portfolio_st_dev <- sqrt(expected_portfolio_variance)

cat(sprintf("The expected portfolio returns standard deviation is %.2f%%\n", expected_portfolio_st_dev * 100))
```
We also see that the expected standard deviation of the portfolio is very close to the simulated one, since we, once again, used the same state probabilities and did a relatively large number of simulations (1000).

```{r}
#We can also confirm this by comparing the theoretical probability of each state, and the relative frequency with which they appear in simulated_states

# Calculate the observed frequency of each state
simulated_states_freq <- table(simulated_states) / length(simulated_states)

# Convert the frequency table into a data frame and ensure it aligns with economic_states
observed_freq <- as.vector(simulated_states_freq[match(economic_states, names(simulated_states_freq))])

# Create the data frame with the correct columns
States_check <- data.frame(
  State = economic_states,
  Theoretical_probability = state_probabilities,
  Observed_frequency = observed_freq
)

# View the first few rows
head(States_check)
```
Now to create the histogram.
```{r}
library(ggplot2)

# Create the histogram of simulated portfolio returns
ggplot(simulated_df, aes(x = Portfolio_Return)) +
  geom_histogram(binwidth = 0.01, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Histogram of Simulated Portfolio Returns",
    x = "Portfolio Return",
    y = "Frequency"
  ) +
  theme_minimal()
```
We can see exactly what was calculated in the table above, how each state happened the exact number of time shown, with the corresponding portfolio returns. 


 This is only a mock portfolio, and the payoffs do not paint a true-to-life picture, but it is still worthy of a few comments for practice.
 With that being said, is quite a conservative portfolio with mean returns close to the returns standard deviation, meaning that we might be better of reallocating our weights or investing in additional assets to secure better returns while maintaining the level of risk (returns standard deviation). 
 Let us assume that this portfolio we have is invested into some European ETFs and government bonds.Good diversification practice would be to invest into the US stock market(could be more volatile but grants higher returns), or perhaps some emerging market ETFs, which tend to have different economic cycles to EU or US economies, thus lowering our risk when EU is in a 'low' state. 
 We could see both of these things in action right now with the code below.
 
```{r}

set.seed(150)

# Number of simulations
n_simulations <- 1000

# Economic states and probabilities
economic_states <- c("High", "Medium", "Low")
state_probabilities <- c(0.30, 0.50, 0.20)

# Asset returns for each state
returns <- data.frame(
  Asset1 = c(0.08, 0.04, -0.02),
  Asset2 = c(0.12, 0.03, -0.05),
  Asset3 = c(0.05, 0.02, 0.01),
  Asset4 = c(0.15, 0.10, -0.10), #ETF following the Nasdaq index for example
  Asset5 = c(-0.04, 0.04, 0.08)  #ETF following some emerging market with difference business cycles
)

# Portfolio weights
weights <- c(0.20, 0.20, 0.20, 0.25, 0.15)

# Simulate 1000 economic states using sample function
simulated_states <- sample(economic_states, n_simulations, replace = TRUE, prob = state_probabilities)

# Generate data for all simulations using lapply
simulated_data <- lapply(simulated_states, function(state) {
  state_index <- which(economic_states == state)
  asset_returns <- returns[state_index,]
  portfolio_return <- sum(asset_returns * weights)
  
  # Return as a data frame
  data.frame(State = state, 
             Asset1_Return = as.numeric(asset_returns[1]), 
             Asset2_Return = as.numeric(asset_returns[2]), 
             Asset3_Return = as.numeric(asset_returns[3]),
             Asset4_Return = as.numeric(asset_returns[4]),
             Asset5_Return = as.numeric(asset_returns[5]),
             Portfolio_Return = portfolio_return, 
             stringsAsFactors = FALSE)
})

# Convert list of data frames to a single data frame
simulated_df <- do.call(rbind, simulated_data)

# View the first few rows
head(simulated_df)


# Let us compute the mean return for each Asset and the portfolio as a whole
A1_mean_return <- mean(simulated_df$Asset1_Return)
A2_mean_return <- mean(simulated_df$Asset2_Return)
A3_mean_return <- mean(simulated_df$Asset3_Return)
A4_mean_return <- mean(simulated_df$Asset4_Return)
A5_mean_return <- mean(simulated_df$Asset5_Return)
Portfolio_mean_return <- mean(simulated_df$Portfolio_Return)

#Now let us print this
cat(sprintf("The mean return of Asset1 is %.2f%%\n", A1_mean_return * 100))
cat(sprintf("The mean return of Asset2 is %.2f%%\n", A2_mean_return * 100))
cat(sprintf("The mean return of Asset3 is %.2f%%\n", A3_mean_return * 100))
cat(sprintf("The mean return of Asset4 is %.2f%%\n", A4_mean_return * 100))
cat(sprintf("The mean return of Asset5 is %.2f%%\n", A5_mean_return * 100))
cat(sprintf("The mean return of the portfolio is %.2f%%\n", Portfolio_mean_return * 100))

#Let's calculate the Asset and portfolio standard deviation now
A1_st_dev <- sd(simulated_df$Asset1_Return)
A2_st_dev <- sd(simulated_df$Asset2_Return)
A3_st_dev <- sd(simulated_df$Asset3_Return)
A4_st_dev <- sd(simulated_df$Asset4_Return)
A5_st_dev <- sd(simulated_df$Asset5_Return)
Portfolio_st_dev <- sd(simulated_df$Portfolio_Return)

#And print this as well,but print some separator in between
cat(rep("*", 39), "\n")
cat(sprintf("Standard deviation of Asset1 returns is %.2f%%\n", A1_st_dev * 100))
cat(sprintf("Standard deviation of Asset2 returns is %.2f%%\n", A2_st_dev * 100))
cat(sprintf("Standard deviation of Asset3 returns is %.2f%%\n", A3_st_dev * 100))
cat(sprintf("Standard deviation of Asset4 returns is %.2f%%\n", A4_st_dev * 100))
cat(sprintf("Standard deviation of Asset5 returns is %.2f%%\n", A5_st_dev * 100))
cat(sprintf("Standard deviation of the portfolio is %.2f%%\n", Portfolio_st_dev * 100))
```
 After adding new Investments and changing the weights, we can see that we have managed to increase the mean return, and even lower the risk, as the portfolio is now more diversified.
