# Introduction {#intro}

Our lectures are going to teach you basic concepts of probability and 
their applications in computational finance. Our approach is based on 
intuition and on understanding concepts by
building them with your own hands.

But how can we build abstract concepts such as probability, random phenomena and chance? 
While probability is a mathematical theory, it gains practical value and an intuitive meaning in
connection with real or conceptual experiments such as, tossing a coin, rolling two dice, playing
roulette, observing the change in the index value of stocks, bonds or in exchange rates, selecting
a random sample of financial transactions in a payment system and observe the number of transactions
with a value below a certain amount.

Many of these experiments we can nowadays do not only physically by actually rolling a pair of dice,
for example but we can use the computer to build a virtual dice and roll it on the computer. We
can build and simulate random phenomena. We can approach our understanding of abstract concepts
by building them with our own hands on the computer. This is the approach to teaching you
probability in this course.

There are many ways to build virtual objects and to run simulations to manipulate 
them. For this we will need a programming language. The language we choose for this
course is R and the integrated development environment RStudio. This is one of the main
languages used in data analysis, statistics, data science and is widely used in industry and 
academia. This introductory chapter will introduce you to this tool in a straightforward and
minimalist way to get us started. As the course develops we will gradually introduce more
and more concepts of R in parallel to our main thread, the introduction to probability.

We will place this introduction to R and RStudio in the context of a little project, which we
then can recycle throughout the course. We will introduce R by learning how to simulate
the rolling a pair of dice, a very concrete example of a process that can be described and 
modeled with concepts from probability and which plays a prominent role in the
teaching of probability. You will learn how to:



This brief introduction to R and RStudio follows very closely the book [@grolemund]. This 
book is an excellent source for a ´ 



Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
