# Conditional Probability

## Introdcution

We begin this chapter with a definition and an example. Here is the definition:

Contingency Table
: A contingency table shows counts of cases on one categorical variable contingent
on the value of another for every combination of both variables.

Contingency tables are useful for calculating **conditional probabilities** as they 
contain all the necessary ingredients for the computation. The concept of
conditional probability will be the main contents of this lecture.

Let us start with an example of a contingency table, which will allow us to repeat at the
same time some R concepts. In our data directory I have stored
the daily prices of the Standard & Poors 500 stock price index and the daily prices
of four stocks, Google, Amazon, Facebook and Apple beginning from January 1st 2015. 
Now we would like to construct 
a contingency table from these data to think about how these individual stocks
move with the overall index. Let us start with Google and you can then do the others
yourself.

First of all let us load the data for the index and write them into a variable called SP500

```{r}
SP500 <- read.csv("data/sp500.csv")
```

We next read the prices of the four individual stocks which are stored in 
one file and write them into the variable `stock_data`

```{r}
stock_data <- read.csv("data/stock_prices.csv")
```

For the contingency table we would like to count the number of days when the
S&P was moving up from one day to the next and how often it was moving down. We would
like to do the same with the Google (Amazon, Facebook, Apple) stock price.

One way to see whether the price moved up or down from one day to the next is to
compute the daily price differences in  the closing price. When this difference is
positive the price made an up movement, when it is negative, it made a down movement (it
could theoretically also not change at all).

There are many ways to implement such a computation in R. An efficient way to do sp one is to use
one of R's internal functions `diff()`. It implements a method for 
computing differences in time series. Consider the following example: Say we have 
a vector of values like
```{r}
val <- c(2,4,7,1,9,10)
```
and we apply the `diff()`function we will get
```{r}
diff(val)
```

Now, of course by taking differences between 4 and 2, 7 and 4, 1 and 7, 9 and 1, ans 10 and 9 we
have to drop the first observation and get a vector which is shorter than val by one component. 
We can
produce a vector of equal length if we add as the first component the value NA, which is
R's symbol of "not available" or of a missing value. Indeed this makes sense because for the
first date if you want we have no value for the change (since we cut off there and don't know
the value from the day before).

Now let us apply this reasoning to the Standard and Poors 500: The data of the index are in a
dataframe, which is a special case of a list. We can add a column by the $ symbol. This
symbol will append a new column with the name we write after $. Let us call this new
column "changes" and compute the changes in the closing price by the `diff()`function. 
Now, since we loose the first value here by taking the first difference, let us 
replace it by NA and from a new column of price differecnes with the first observation
replace by NA. Thus we type:

```{r}
SP500$changes <- c(NA, diff(SP500$close))
```

Lets do the same for the Google stock price. If you inspect the data by using the `View()` function
you will see that the price series of our different stocks 
are stacked upon each other. How can we filter just the
Google Price? Here we can make use of Rs system of selecting values from lists (since the
stock prices are a dataframe, a special list). We type:

```{r}
google <- stock_data[stock_data$symbol == "GOOGL", ]
```

Let us explain: We can filter the rows which have the symbol "GOOGL" by telling R to select
the rows - rows are the first entries in stock_data[rows,columns] after the [ bracket.
For example, if we would type 

```{r}
stock_data[1:10,1:6]
```
we get row 1 to 10 and column 1 to 6. Now this syntax can also be used by 
specifying conditions. The
condition we use is that the values of the rows we look for are all rows, where
the value of the column stock_data$symbol is equal to the string "GOOGL" and other than this
we select all columns. To express this equality we need the symbol `==` (and not `=` which
works as an assignment operator in R, similar to `<-`).

This is expressed as:

```{r}
google <- stock_data[stock_data$symbol == "GOOGL", ]
```

Now as before with the Standard and Poors 500 Index we can add a column recording the changes
in the closing price

```{r}
google$changes <- c(NA, diff(google$close))
```

Now we add two new columns to the SP500 dataframe, by assigning a logical operation. Whenever
the price movment went up the first difference will be positive. R checks this condition for
every row and writes TRUE if the condition is fulfills and FALSE otherwise. An analogous
operation can be done for the up movements of Google's stock price.

```{r}
SP500$Up <- SP500$changes > 0
google$Up <- google$changes > 0
```

We can now compute a contingency table of both movements. R has a function, called `table()` 
which does this job. 

```{r}
table(SP500$Up, google$Up)
```

The contingency table shows counts of Up and Down movements of the Stadard and Poors index closing
price and the price of Google. In all the 1699 observations, there are 556 days when
the Index moves down and the Price of Google also moves down. There are only 218 days when the
Index moves down but Google moves in the other direction Up. There are even more days when
the Index and the Google stock price move up together. Using a relative frequency interpretation
of probability we can transform these counts into probabilities by dividing the number by the
number of total observations 1699. Let's do this:

```{r}
table(SP500$Up, google$Up)/1699
```

These are the probabilities of joint moves or cross moves. 

## Marginal, joint and conditional probability

Marginal probabilty
:The **marginal probability** is the unconditional probability on an event. This 
probability is not conditioned on any other event. In terms of notation, we
denote the marginal probabilities of two arbitrary events $A$ and $B$ by
$P(A)$ and $P(B)$ respectively.

Joint probability
: The **joint probability** is the probability of simultaneous events. This is
the probability of the intersection of two or more events. If $A$ and $B$ are
arbitrary events, we write $P(A \cap B)$.

Conditional probability
: The **conditional probability** is the probability of an event *given* that some other 
event has occurred. The information from the event that has occured can influence
the probability of the original event. The following notation is used for
conditional probabilities: For two events $A$ and $B$, the conditional probability of
$A$ given $B$ is denoted by $P(A\,|\,B)$ and the conditional probability of $B$ given
$A$ is denoted as $P(B\, |\, A)$. In general these two conditional probabilities are
not the same.

The probability table of joint moves, we have computed before can help us to illustrate 
these concepts and state a formula for the conditional probability.

|          | Google-Down | Google-Up | Total |
|----------|:-----------:|:---------:|------:|
|SP500-Down| 0.33        | 0.13      | 0.46  |
|SP500-Up  | 0.13        | 0.41      | 0.54  |
|Total     | 0.46        | 0.54      | 1.00  |

The **marginal probability** that the SP500 will make an Up move is 
$P(\text{SP500-Up} \cap \text{Google-Down}) + P(\text{SP500-Up} \cap \text{Google-up}) = P(\text{SP500-Up})$, which is 0.54, given our table. It is the sum
of the two **joint probabilities** $P(\text{SP500-Up} \cap \text{Google-Down})$ and 
$P(\text{SP500-Up} \cap \text{Google-Up})$. To compute the conditional probability
that the stock price of Google will make an up move given that the SP500 makes an 
up move we have to compute $P(\text{Google-Up} | \text{SP500-Up})$. The probability
that the Google price went up and the SP500 went up is $P(\text{Google-up} \cap \text{SP500-Up})$.
But given I already know that by the condition SP500 went up, I am in the second line of
my table and we know that the first line, SP500 went down - is ruled out by the
condition. We thus have now a *new sample space* and we must 
adjust the probabilities proportionally
to be consistent with the rules of probability. We therefore have to divide by
$P(\text{SP500-Up})$ to get the formula
$P(\text{Google-Up} | \text{SP500-Up}) = \frac{P(\text{Google-up} \cap \text{SP500-Up})}{P(\text{SP500-Up})}$ which is $0.41/0.54$ or $0.76$.
To check whether we have made the correct renormalisation you can ask yourself: Given the
rules of probability, what is $P(\text{Google-Down} | \text{SP500-Up}))$ ? It should be
$1 - 0.76$ or 0.24. Applying the formula for conditional probability we would compute
$0.13/0.54$ which is indeed 0.24.

Another way to think about this formula can perhaps help to foster our intuition if we
look at this problem through the frequency interpretation of probability:

We would like to
find the conditional probability that Google makes an Up move given that 
SP500 makes an up move. Suppose the daily price moves are a conceptual 
random experiment with the outcomes that the
price can either move up or down and we have two prices SP500 and Google, so our
sample space has four possible outcomes:
${\cal S} = \{ UU, UD, DU, DD \}$ where the first character if it is U means SP500-Up 
and the second if it is U means Google-Up and so on for all possible combinations.
We repeat this 
experiment 1699 times and keep a notebook where at every repetition we write down the
number of the experiment, the outcome and then we note whether both Google and the SP500
went up.

| notebook line | outcome             | SP500-Up and Google-Up ?|
|:-------------:|:-------------------:|:-----------------------:|
|1              |SP500-Up, Google-Down| No                      |
|2              |SP500-Down, Google-Up| No                      |
|3              |SP500-Up, Google-Up  | Yes                     |
|.              |.                    |.                        |
|.              |.                    |.                        |
|.              |.                    |.                        |
|1699           |SP500-Up, Google-Up  |Yes                      |

To get the conditional probability of a Google-Up given an SP500-Up we have to count
the Google-Ups **among the notebook lines** where the SP500 has an Up move and divide by
the number of these SP500-Up lines.

Let's do this in R. To create the notebook we make a dataframe where one column is the Up column
of the SP500 data and the second columns is the Up column of the Google data.

```{r}
joint_ups <- data.frame(SP500Up = SP500$Up, GoogleUp = google$Up)
```

Now we apply what we h ave already learned about referencing values in a dataframe in R to
select all the number in our fictitious notebook where the SP500 moves Up:

```{r}
joint_ups[joint_ups$SP500Up == TRUE, ]
```

To count the number of rows we can use the `dim()` function.

```{r}
dim(joint_ups[joint_ups$SP500Up == TRUE, ])
```

It says that the number of notebook lines where the SP500 makes an Up move are 926. Since the
first line is an NA we have to subtract this line, so we have 925 lines with actual Ups in the
SP500.

Now, let us count the Ups in the Google occurring simultaneously with the SP500 ups

```{r}
joint_ups$simultaneous = (joint_ups$SP500Up & joint_ups$GoogleUp)
```

Now we can make use of R's precedence rules to count the simultaneous up moves

```{r}
sum(joint_ups$simultaneous, na.rm = T)
```

Now we compute $702/926$ which gives us 0.76 as we have already verified before.

Note that for conditional probabilities we have for two events $A$ and $B$, that 
$P(A|B) \neq P(B|A)$. To see this assume that $P(A) \neq P(B)$ and $P(A) \neq 0$ and 
$P(B)\neq 0$. We then get:
$P(A|B) = \frac{P(A\cap B)}{P(B)} = \frac{P(B \cap A)}{P(B}$, since $P(A \cap B) = P(B \cap A)$.
It follows that $\frac{P(B \cap A)}{P(B} \neq \frac{P(B \cap A)}{P(A} = P(B|A)$, since we
has assumed that $P(A) \neq P(B)$. Therefore $P(A|B) \neq P(B|A)$.

Try it with the stock price example above.

The notion of conditional probability also gives us a new insight into the notion of
independence, we discussed earlier. Remember that two events $A$ and $B$ are independent
if $P(A \cap B) = P(B \cap A) = P(A) \cdot P(B)$. If we combine this rule with the 
concept of conditional probability, we see that if two events $A$ and $B$ are independent, then
$P(A|B) = \frac{P(A \cap B)}{P(B)} = \frac{P(A) \cdot P(B)}{P(B)} = P(A)$ and
$P(B|A) = \frac{P(B \cap A)}{P(A)} = \frac{P(A) \cdot P(B)}{P(A)} = P(B)$.

This formula says that if two events are independent the probability of $A$ is 
not influenced by the event $B$ occuring and the probability of event $B$ is not
influenced by the event $A$ occurring.

## Birthdays and Blockchains continued

Let us go back to the problem we discussed in lecture 1 where we had $n$ people in a class. How
much would you bet that there is nobody in the room who shares the same birthday?

We would like to compute the probability of the event that at least two people have the
same birthday, $P(A)$. We compute this via the simpler problem that all $n$ students have
a different birthday and by using the rules of probability. 

Since the union of an event $A$ and
its complement $A^C$ is the whole sample space ${\cal S} = A \cup A^C$ and $A \cap A^C = \emptyset$, we have
$P({\cal S}) = P(A) + P(A^C)$. Now since $P({\cal S}) = 1$, this means we can 
compute the probability we are looking for as: $1 - P(A^C)$.

In lecture 1 we have already found that for $n$ students, assuming that every birthday is
equally likely, we have $365^n$ possible outcomes for all students having a different birthday
and this can happen in $365\cdot 364\cdot 363,\ldots, (365-n+1)$ ways. Therefore we get
$1 - P(A^C) = 1 - \frac{365\cdot 364\cdot 363,\ldots, (365-n+1)}{365^n}$

Now we can use this formula to write a function in R which gives us the probability as a
function of the number of people in a room.

```{r}
birthday_collision <- function(n){
  
  1 - prod(365:(365-n+1))/365^n
  
}
```

In terms of R this should look all familiar to you: You know already how to write functions, we
use the R function `prod()` giving the product of an arbitrary finite number of factors. We did
not explicitly introduce this function but you can look up its details using the help facilities of R. The colon operator `:` you have alsoe learned already. In this example it would create
a vector of numbers from 365 to 356, which is then taken as the input to the `prod()` function. That's it. 

If there are 10 people:

```{r}
birthday_collision(10)
```

If there are 23 people in the room this probability is already

```{r}
birthday_collision(23)
```

more than 50 %. This is quite stunning because intuitively perhaps most of you 
including myself would assume that this probability is much smaller. This is fun, but 
what does this example have to do with computational Finance?

In lecture 1 we have explained that there is a common context with the birthday problem 
and collisions of a hash function. Lets work out the collision probabilities for the
SHA-256 hash function. In the analogy of the birthday problem and hash-collsions the
number $n$ of students in class are the inputs and their birthdays can be considered the
analogy to the hash values. When two people share the same birthday this is analogous to 
a hash collision.

A hash with 256 bits can store $2^{256}$ combinations of output values. For values
of this size we cannot make precise computations using the exact formula. This is
because the floatimg point representation of numbers will not allow us to distinguish
very large numbers. We need to work with approximations. It can be shown that the
expression:
$\frac{M - 1}{M} \cdot \frac{M-2}{M} \cdot \frac{M-3}{M} \ldots \frac{M-(n-2)}{M} \cdot \frac{M-(n-1)}{M}$ is approximately equal to $\exp(\frac{-n(n-1)}{2 M})$ and that this
approximation is good in the sense that the approximation error gets smaller as $M$ gets larger
and eventually vanishes as $M$ goes to infinity. We do not explain this calculus based argument 
in detail here but those of you who are interested will find lost of references by a simple
web search.

If you take as an output range a hash function with 32 bits you can use this formula to convince
yourself that the number of hashes needed to get a collision probability of 50 \% is at
77163 hashes. This is not good enough for cryptographic security. 

If we want to get the numerics for such lare ranges as 256 bit hash functions we even need
another approximation. If the exponent is very small, i.e. M very large compared to n we can use
the further approximation $1-\exp(-X) \approx X$. This means that $\frac{n(n-1)}{2 M}$ is already 
good enough for very large $M$ and is $n$ is relatively small then this can be again
simplified to $\frac{n^2}{2 M}$. Taking this approximation we see that to make the 
probability that no collision occurs $99.8 \%$ you need with a 256 bit hash function you can
hash $n = \sqrt{2 \cdot M \cdot 0.998}$ values before that occurs. This is a number of
the magnitude $4.807505 \cdot 10^38$. At current hardware capacity this is several times the
age of the universe. So for practical purposes this can be considered safe.

Finally this example shows that while the computer is a powerful machine at very large numbers
brute force programming won't help because you hit the limits of numerical computation. You need
to think about how to deal with such situations. 

# Bayes' Rule

The definition of conditional probabilities has an important implication about the relation 
between the probabilities of events, which is known as *Bayes' rule*. Lets say we have two
events $A$ and $B$, then we know from the definition of conditional probability that
$P(B|A) = \frac{P(B \cap A)}{P(A)}$. Since $P(B \cap A) = P(A \cap B)$ we have
$P(B|A) = \frac{P(B \cap A)}{P(A)} = \frac{P(A \cap B)}{P(A)}$ which is by definition
equal to $\frac{P(A|B) \cdot P(B)}{P(A}$ provided that $P(A) \neq 0$. The terminology
calls $P(B)$ the *prior probability* of $B$ and $P(B|A)$ is called the *posterior probability*.

One way to think about the relation
$P(B|A) = \frac{P(A|B) \cdot P(B)}{P(A}$ is to think about the event $B$ as some hypothesis and
the event $A$ as data or information.
## R-Concepts

Pack the R concepts into the Apple and Stock market example.

Example strutured finance and violations of independence

## Project