# Conditional Probability

## Introdcution

We begin this lecture with a definition and an example. Here is the definition:

Contingency Table
: A contingency table shows counts of cases on one categorical variable contingent
on the value of another for every combination of both variables.

Contingency tables are useful for calculating **conditional probabilities** as they
contain all the necessary ingredients for the computation. The concept of
conditional probability and its applications
will be the main contents of this lecture.

Let us start with an example of a **contingency table** within the context
of stock price moves, we discussed already to some
degree in lecture 2. It gives me also an opportunity to point you to one
powerful R package, which will allow you to access financial data, quite easily: The
`tidyquant-package` (https://business-science.github.io/tidyquant/).

Remember from lecture 1 that we can install add-on packages for R, which extend the
functionality of the system by new functions. A packages has to be installed first and then
loaded into working memory. To install tidy-quant let us first use the install command:
`install.packages("tidyquant")` 
and then load the package by:
```{r}
library(tidyquant)
```

We do not go into the tidyquant package in any detail. If you are interested
check out the github page cited above. For now we only show the
use of one of its core functions which allows you to load financial data directly
from the internet. 

This function is called `tq_get()`. Let's see how it works. We first
retrieve the S&P500 stock market index. To do this we need to know the ticker symbol, which is
`^GSPC`. This has to be passed to the `tq_get()` function and tha's it. Let us save these 
data into an R object, which I call `sp500`and jook at the first entries
```{r load-sp-500}
sp500 <- tq_get("^GSPC")
head(sp500, n = 5)
```
We see that the series starts March 2011.
Now let's retrieve the stock price data of 
Apple beginning at 2011-01-03, the same day as the index data.
```{r read-others}
apple <- tq_get("AAPL", from = " 2011-01-03")
```
You can give a beginning date to the argument `from` in the `tq_get()` function (you can guess 
that you can also give an end data `to` as an argument to `tq_get`)

For the contingency table we would like to count the number of days when the
S&P was moving up from one day to the next and how often it was moving down. We would
like to do the same with the Apple stock price.

One way to see whether the price moved up or down from one day to the next is to
compute the daily price differences in  the closing price. When this difference is
positive the price made an up movement, when it is negative, it made a down movement (it
could theoretically also not change at all).

There are many ways to implement such a computation in R. Let me remind
you of the `diff()` function, we already used in lecture 2. It implements a method for
computing differences in time series. Consider the following example: Say we have
a vector of values like
```{r}
val <- c(2,4,7,1,9,10)
```
and we apply the `diff()`function we will get
```{r}
diff(val)
```

Now, of course by taking differences between 4 and 2, 7 and 4, 1 and 7, 9 and 1, ans 10 and 9 we
have to drop the first observation and get a vector which is shorter than val by one component.
We can
produce a vector of equal length if we add as the first component the value NA, which is
R's symbol of "not available" or of a missing value. Indeed this makes sense because for the
first date if you want we have no value for the change (since we cut off there and don't know
the value from the day before).

Now let us apply this reasoning to the Standard and Poors 500: The data of the index are in a
data frame. We can add a column to this data frame by the $ symbol followed by the
column name. This
symbol will append a new column with the name we write after $. Let us call this new
column "changes" and compute the changes in the closing price by the `diff()`function.
Now, since we loose the first value here by taking the first difference, let us
replace it by NA and from a new column of price differences with the first observation
replace by NA. Thus we type:

```{r changes-sp500}
sp500$changes <- c(NA, diff(sp500$close))
```
Lets do the same for the Aoogle stock price. 
```{r}
apple$changes <- c(NA, diff(apple$close))
```
We want to count the up moves and the down moves. Since there are only these two outcomes
it is sufficient if we add a logical column to both dataframes, checking the condition that
the change variable is larger than zero.
```{r}
sp500$up <- (sp500$changes > 0)
apple$up <- (apple$changes > 0)
```
We can now tabulate the joint up and down moves using the `table()`function
```{r table}
table(sp500$up, apple$up)
```
Now this looks already good and like a contingency table. The problem is we do not see
exactly what is what. So let is check: Using R's coercion rules we can count these moves by applying the `sum()`function to each
column. Since sum requires values of type double of integer, R will coerce `TRUE`to 1 and
`FALSE`to zero. Note that, since the first value in each variable is NA we have to give the
argument `na.rm`to the sum function. 
```{r}
sum(sp500$up, na.rm = T)
sum(apple$up, na.rm = T)
```
This says that sp500 moves up 1516 times and apple moves up 1451 times. This means sp500
is in the rows and apple is in the columns in the output of the `table()`function. So let us 
write this down more neatly:


|          | Apple-Down | Apple-Up | Total |
|----------|:-----------:|:---------:|------:|
|SP500-Down|  858        |  378      | 1236  |
|SP500-Up  |  443        | 1073      | 1516  |
|Total     | 1301        | 1451      | 2752  |

This table says that there are 1073 cases where the SP500 index and the Apple stock price move 
up together, there are 858 cases where they fall together and then there are 443 cases and
378 cases abd 443 cases respectively where the variables move in opposite directions.

## Marginal, joint and conditional probability

Marginal probabilty
:The **marginal probability** is the unconditional probability on an event. This
probability is not conditioned on any other event. In terms of notation, we
denote the marginal probabilities of two arbitrary events $A$ and $B$ by
$P(A)$ and $P(B)$ respectively.

Joint probability
: The **joint probability** is the probability of simultaneous events. This is
the probability of the intersection of two or more events. If $A$ and $B$ are
arbitrary events, we write $P(A \cap B)$.

Conditional probability
: The **conditional probability** is the probability of an event *given* that some other
event has occurred. The information from the event that has occurred can influence
the probability of the original event. The following notation is used for
conditional probabilities: For two events $A$ and $B$, the conditional probability of
$A$ given $B$ is denoted by $P(A\,|\,B)$ and the conditional probability of $B$ given
$A$ is denoted as $P(B\, |\, A)$. In general these two conditional probabilities are
not the same.

The table of joint moves, we have computed before can help us to illustrate
these concepts using the frequency interpretation of probability: We divide 
each entry in our table by the total number of cases

|          | Apple -Down | Apple -Up | Total |
|----------|:-----------:|:---------:|------:|
|SP500-Down| 0.31        | 0.14      | 0.45  |
|SP500-Up  | 0.16        | 0.39      | 0.55  |
|Total     | 0.47        | 0.53      | 1.00  |

The **marginal probability** that the SP500 will make an Up move is
$P(\text{SP500-Up} \cap \text{Google-Down}) + P(\text{SP500-Up} \cap \text{Google-up}) = P(\text{SP500-Up})$, which is 0.55, given our table. It is the sum
of the two **joint probabilities** $P(\text{SP500-Up} \cap \text{Apple-Down})$ and
$P(\text{SP500-Up} \cap \text{Apple-Up})$. To compute the conditional probability
that the stock price of Google will make an up move given that the SP500 makes an
up move we have to compute $P(\text{Apple-Up} | \text{SP500-Up})$. The probability
that the Google price went up and the SP500 went up is $P(\text{Apple-up} \cap \text{SP500-Up})$.
But given I already know that by the condition SP500 went up, I am in the second line of
my table and we know that the first line, SP500 went down - is ruled out by the
condition. We thus have now a *new sample space* and we must
adjust the probabilities proportional
to be consistent with the rules of probability. We therefore have to divide by
$P(\text{SP500-Up})$ to get the formula
$P(\text{Apple-Up} | \text{SP500-Up}) = \frac{P(\text{Apple-up} \cap \text{SP500-Up})}{P(\text{SP500-Up})}$ which is $0.39/0.55$ or $0.71$.

To check whether we have made the correct re-normalisation you can ask yourself: Given the
rules of probability, what is $P(\text{Apple-Down} | \text{SP500-Up}))$ ? It should be
$1 - 0.71$ or 0.29. Applying the formula for conditional probability we would compute
$0.16/0.55$ which is indeed 0.29.

Another way to think about this formula can perhaps help to foster our intuition if we
look at this problem through the frequency interpretation of probability:

We would like to
find the conditional probability that Apple makes an up move given that
SP500 makes an up move. Suppose the daily price moves are a conceptual
random experiment with the outcomes that the
price can either move up or down and we have two prices SP500 and Apple, so our
sample space has four possible outcomes:
${\cal S} = \{ UU, UD, DU, DD \}$ where the first character if it is U means SP500-Up
and the second if it is U means Apple-Up and so on for all possible combinations.
We repeat this
experiment 2752 times and keep a notebook where at every repetition we write down the
number of the experiment, the outcome and then we note whether both Apple and the SP500
went up.

| notebook line | outcome             | SP500-Up and Google-Up ?|
|:-------------:|:-------------------:|:-----------------------:|
|1              |SP500-Up, Google-Down| No                      |
|2              |SP500-Down, Google-Up| No                      |
|3              |SP500-Up, Google-Up  | Yes                     |
|.              |.                    |.                        |
|.              |.                    |.                        |
|.              |.                    |.                        |
|1699           |SP500-Up, Google-Up  |Yes                      |

To get the conditional probability of a Apple-Up given an SP500-Up we have to count
the Google-Ups **among the notebook lines** where the SP500 has an Up move and divide by
the number of these SP500-Up lines.

Let's do this in R. To create the notebook we make a data frame where one column is the Up column
of the SP500 data and the second columns is the Up column of the Google data.

```{r}
joint_ups <- data.frame(SP500Up = sp500$up, AppleUp = apple$up)
```

Now we apply what we h ave already learned about referencing values in a data frame in R to
select all the number in our fictitious notebook where the SP500 moves Up:

```{r}
joint_ups[joint_ups$SP500Up == TRUE, ]
```

To count the number of rows we can use the `dim()` function.

```{r}
dim(joint_ups[joint_ups$SP500Up == TRUE, ])
```

It says that the number of notebook lines where the SP500 makes an Up move are 1518. Since the
first line is an NA we have to subtract this line, so we have 925 lines with actual Ups in the
SP500.

Now, let us count the Ups in the Google occurring simultaneously with the SP500 ups

```{r}
joint_ups$simultaneous = (joint_ups$SP500Up & joint_ups$AppleUp)
```

Now we can make use of R's precedence rules to count the simultaneous up moves

```{r}
sum(joint_ups$simultaneous, na.rm = T)
```

Now we compute $702/926$ which gives us 0.76 as we have already verified before.

Note that for conditional probabilities we have for two events $A$ and $B$, that
$P(A|B) \neq P(B|A)$. To see this assume that $P(A) \neq P(B)$ and $P(A) \neq 0$ and
$P(B)\neq 0$. We then get:
$P(A|B) = \frac{P(A\cap B)}{P(B)} = \frac{P(B \cap A)}{P(B}$, since $P(A \cap B) = P(B \cap A)$.
It follows that $\frac{P(B \cap A)}{P(B} \neq \frac{P(B \cap A)}{P(A} = P(B|A)$, since we
has assumed that $P(A) \neq P(B)$. Therefore $P(A|B) \neq P(B|A)$.

Try it with the stock price example above.

The notion of conditional probability also gives us a new insight into the notion of
independence, we discussed earlier. Remember that two events $A$ and $B$ are independent
if $P(A \cap B) = P(B \cap A) = P(A) \cdot P(B)$. If we combine this rule with the
concept of conditional probability, we see that if two events $A$ and $B$ are independent, then
$P(A|B) = \frac{P(A \cap B)}{P(B)} = \frac{P(A) \cdot P(B)}{P(B)} = P(A)$ and
$P(B|A) = \frac{P(B \cap A)}{P(A)} = \frac{P(A) \cdot P(B)}{P(A)} = P(B)$.

This formula says that if two events are independent the probability of $A$ is
not influenced by the event $B$ occurring and the probability of event $B$ is not
influenced by the event $A$ occurring.

## Example: Independence and the Financial Crisis of 2007-2008


## Bayes' Rule

The definition of conditional probabilities has an important implication about the relation
between the probabilities of events, which is known as *Bayes' rule*. Lets say we have two
events $A$ and $B$, then we know from the definition of conditional probability that
$P(B|A) = \frac{P(B \cap A)}{P(A)}$. Since $P(B \cap A) = P(A \cap B)$ we have
$P(B|A) = \frac{P(B \cap A)}{P(A)} = \frac{P(A \cap B)}{P(A)}$ which is by definition
equal to $\frac{P(A|B) \cdot P(B)}{P(A}$ provided that $P(A) \neq 0$. The terminology
calls $P(B)$ the *prior probability* of $B$ and $P(B|A)$ is called the *posterior probability*.

One way to think about the relation
$P(B|A) = \frac{P(A|B) \cdot P(B)}{P(A}$ is to think about the event $B$ as some hypothesis and
the event $A$ as data or information.
## R-Concepts

Example structured finance and violations of independence

## Project