
---
title: "Introduction to Probability and R"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  beamer:
    theme: metropolis
    slide-level: 2
    toc: true
    incremental: false
execute:
  echo: true
  warning: false
  message: false
---

# Preface

> "What I hear, I forget; What I see, I remember; What I do, I understand"  
> — Confucius, 551-479 BC

---

## What are these lecture notes about?

These lecture notes support the course **An Introduction to Probability - with Applications to Computational Finance using R**. The course introduces essential probability concepts that every finance practitioner must understand.

Key highlights:
- Hands-on learning approach.
- Leverages **R** for simulations and visualizations.
- Connects theory with real-world applications.

---

## Key Learning Objectives

1. Understand foundational probability concepts.  
2. Apply probability principles to solve computational finance problems.  
3. Gain hands-on experience using R for probability and finance applications.  
4. Develop proficiency in interpreting and visualizing probabilistic data.

---

## Building Intuition through Simulation

- Probability concepts gain value when connected to real or conceptual experiments:
  - Stock price changes.
  - Portfolio value fluctuations.
  - Credit risk evaluations.

- Simulations allow us to:
  - Construct and manipulate random phenomena.
  - Model financial risks and explore future scenarios.

---

## Example: Simulating Stock Returns in R (Code)

```{r}
set.seed(123)  # Ensure reproducibility
n <- 100       # Number of days
daily_returns <- rnorm(n, mean = 0.001, sd = 0.02)

price <- cumprod(1 + daily_returns) * 100  # Starting price = 100

plot(price, type = "l", col = "blue", lwd = 2,
     main = "Simulated Stock Price",
     xlab = "Days", ylab = "Price")
```

---

## Example: Simulating Stock Returns in R (Visualization)

```{r fig.width=7, fig.height=4, out.width='80%'}
# Ensure the plot is scaled properly
plot(price, type = "l", col = "blue", lwd = 2,
     main = "Simulated Stock Price",
     xlab = "Days", ylab = "Price")
```

- **Highlights**:
  - Influence of randomness on stock prices.
  - R’s utility for simulations and visualization.

---

## Using R and RStudio

### Installing R
1. Download from [CRAN](http://cran.r-project.org).  
2. Choose your operating system (Windows, Mac, Linux).  
3. Install precompiled binaries (easiest method).  

### Installing RStudio
1. Download RStudio Desktop (free) from [RStudio website](https://www.rstudio.com/products/rstudio/).  
2. Ensure R is installed first.

---

## Large Language Models (LLMs) for Learning

- LLMs like ChatGPT can assist with:
  - Clarifying concepts.
  - Generating examples.
  - Debugging R code.
  - Creating practice exercises.

---

# Acknowledgments

I thank **Branko Urošević** for entrusting me with developing this course. I am also grateful to former students for their feedback and enthusiasm.

---

# References

- **Feller, W. (1968)**. *An Introduction to Probability Theory and Its Applications, Vol.1*.  
- **Schmedders, K. (Coursera)**. *An Intuitive Introduction to Probability*.  
- **Grolemund, G. (2014)**. *Hands-on Programming with R*.  
- **Luenberger, D. (2009)**. *Investment Science*.  
- **Diaconis, P., & Skyrms, B. (2019)**. *10 Great Ideas About Chance*.

