
---
title: "First Probability Ideas and First Steps in R"
author: "Your Name"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  beamer:
    theme: metropolis
    slide-level: 2
    toc: true
    incremental: false
execute:
  echo: true
  warning: false
  message: false
---

# Finance and Uncertainty

- Finance inherently involves **uncertainty** and **risk**.
- Core questions in Finance:
  - How to allocate and price money over time?
  - How to quantify and manage uncertainty?
- Examples:
  - **Saving**: Deferring consumption today for future benefits.
  - **Borrowing**: Investing now, with repayment tied to uncertain future revenues.

---

## Role of Probability

- **Risk and uncertainty** are unavoidable in Finance.
- Probability provides the tools to:
  - Quantify uncertainty.
  - Analyze and manage risks.
- A cornerstone of **financial decision-making**.

---

## Historical Context

- Probability theory began as a mathematical discipline.
- Origins trace back to **16th and 17th century Europe**:
  - Debates about gambling.
  - Involvement of great minds: Cardano, Pascal, Fermat, Bernoulli.
- Earlier human awareness of chance:
  - Gambling traditions.
  - Veneration of the **goddess of chance**.

---

# First Steps: Coin Tossing

- Tossing a coin:
  - **Classical probability**: Equal chance for heads and tails.
  - A simple chance process people have used throughout history.
- Link to Finance:
  - Simulates stock behavior with equal chances of price increase or decrease.

---

## Simulating a Coin Toss in R (Code)

```{r}
set.seed(42)  # Ensures reproducibility
n_tosses <- 100
results <- sample(c("Heads", "Tails"), size = n_tosses, replace = TRUE)

# Print the first 10 results
print(results[1:10])
```

---

## Visualizing Coin Toss Results (Code and Plot)

```{r fig.width=7, fig.height=4, out.width="80%"}
# Create a bar plot of the outcomes
barplot(table(results), col = c("blue", "red"),
        main = "Coin Toss Outcomes",
        xlab = "Outcome", ylab = "Frequency")
```

---

# Key Takeaways

1. Probability is essential for quantifying and managing uncertainty in Finance.
2. Simulations like coin tossing provide a simple way to explore probabilistic concepts.
3. R is a powerful tool for performing these simulations and visualizations.

