---
title: "Investment game: Project 3"
output: html_notebook
author: Martin Summer
date: 17.12.2021
---

In this lecture, unlike in the last,  we have introduced many new probability concepts but
only a few new things from R. In this project we bring both together again in a context
of an investment game. 

Suppose you participate in an investment game. You get a starting wealth of
1000 Euro. You are judged at the end of the game by how much you have been
able to grow this initial wealth.

There is a referee in the game who can choose among two different coins
which he may flip: Green coins and red coins. Green coins are
physically designed such that they land heads with a probability of 0.7 and 
tails with 0.3. Red coins
are physically designed the opposite way, 0.3 chance of landing on heads 0.7 chance landing
on tails.

The game consists of 1000 coin flips. At the
beginning of the game's referee fills a large bucket with a large number of coins, but
it is unknown how many of them are green and how many are red.
He then randomly draws a single coin 
from the bucket. He uses this coin for all 50 flips in that round making sure 
the color is hidden for you and all the other participants.

Before each flip of the coin the ownership to the flip is auctioned off. 
Once the owner of the flip is set, the referee
flips the coin. If it land heads it pays out 2 to the owner. If it lands tails the owner
gets nothing. After each round the referee reveals the result of the flip to
the participants and opens bidding for the next flip.

1. Assume that you know the color of the coin and  the information about its 
probabilities of landing heads of tail: What is the maximum price you should 
pay for the green coin and what is the
maximum price you should pay for the red coin?

Answer: Upon a single flip each coin will either pay 2 or 0. But over a number of
50 flips the green coin will pay on average:
```{r}
0.7*2 + 0.3*0
```
and the red coin will pay on average:
```{r}
0.3*2 + 0.7*0
```
If you buy a green coin above 1.40 you will make a loss in the round, and if you buy a
red coin above 0.6 you will also make a loss in the round.

2. To get an accurate picture of what each flip is worth you need to find out whether the
referee is using a red or a green coin. But this information is not available to you. But you
might form a belief or assign an assumed probability, perhaps based on your counting of
red and green coins in the bucket. Combining this probability with the probability of the
coin landing heads or tails you can make an estimated of the worth of each flip. This estimate
will be somewhere between the maximum value of a red and the maximum value of a 
green coin, depending on how much weight you give to each color. Suppose now you have
examined the bucket and found that 9 out of 10 coins in there were green. How much would
you be willing to maximally pay given this assumption:

Answer:

```{r}
0.9*0.7*2 + 0.1*0.3*2
```

3. Now assume you start with the belief of 90 % that the coin is green and the referee has in fact
drawn a red coin. Create a simulation of the investment game, where the rounds are shown 
in a data frame. Each row of the data frame shows one flip of the coin. The columns
of the data frame should show the actual outcome of  the flip of a red coin, the cumulative
sum of wins, the number of total flips, the probability of the outcome given 
the assumption that the coin is green, the prior probability you have gained from examining the
bucket, the complementary probability of the outcome given the assumption that the coin is green,
and the complementary probability to the prior probability, 