---
title: "Project 2: Can the EBA data be trusted?"
output: html_notebook
author: Martin Summer
date: 8.1.2022
---

The European Banking Authority EBA conducts a biannual stress testing exercise for the
biggest European banks every two years. In the stress test bank exposure data are collected
and then particular stress scenarios are applied to the reported data. The goal is to
find out whether the banks would have enough capital to withstand an adverse scenario. EBA
makes a huge effort to publish the data that are the base of the stress test on its web site:
https://www.eba.europa.eu/risk-analysis-and-data/eu-wide-stress-testing

In this project you are asked to use the knowledge you learned about 
empirical probabilities, R and Benford's law to check whether the EBA stress testing
data show the distribution of leading digits, you would expect from Benford's law.

1. Go to the EBA website https://www.eba.europa.eu/risk-analysis-and-data/eu-wide-stress-testing
and download the 
file Credit Risk IRB (https://www.eba.europa.eu/assets/st21/full_database/TRA_CRE_IRB.csv).
You can do so by downloading the file, storing it locally and then read it into R or you can
read it directly from the web.

2. We want to study the distribution of leading digits in the Amounts reported in the EBA
file of credit risk exposures. Now the csv file contains many different data that all somehow
belong to the EBA stress test. We would not like to check all data but only
exposure data. We need first to filter the data to make sure we have a meaningful collection
of exposure data for all the banks. The description of the data and the data dimensions is 
in the files `Metadata_TR.xlsx` and `Data_Dictionary.xlsx`. You are welcome to study these
data in detail. It will probably need more time than you have available. They are also quite
complicated. Since the aim of this project is not directly to understand the eba-data but to
work with your R-concepts and probability concepts, let me guide you here how to filter 
these data in 10 sequential steps. Note that this sequencing is for didactical reasons 
only and for the 
purpose not to loose oversight. With routine and experience all these steps can be done in
one go as well:
   (a) Extract all variables names, using the `names()`function.
   (b) Select all rows where the Scenario variable has value 1. Note
       that the symbol you need in the R syntax for equal is `==`, the syntax is therefore 
       `Scenario == 1`. You might check out the R-help entry `Comparison` for further details.
   (c) From the resulting data-frame select all rows where the Country variable is not equal to 0.        (hint: The not equal operator in the R syntax is `!=`). If you look into the Metadata-File
       you will see that 0 are all the aggregate exposures not broken down by country. Excluding 
       these will give us country exposures.
   (d) From the resulting data frame select all rows where the Portfolio variable has value 1 or         2.These codes describe the accounting rules under which the exposure values are reported, 
       internal rating based (IRB) or standard approach (SA). As a hint you can use R's subset
       operator `%in%` here so `Portfolio %in% c(1,2)` written with the approprate subsetting rule
       will select all rows where the Porfolio variable has value 1 or 2.
   (e) From the resulting data frame choose all the rows where the Exposure variable is not 0.This
       gives again disaggregated numbers.
   (f) From the resulting data frame choose all the rows where the Status variable has value 0.
   (g) From the resulting data frame choose all the rows where the IFRS9_Stages variable has value        1,2, or 3.
   (h) From the resulting data frame choose all the rows where the CR_guarantees variable is 0
   (i) From the resulting data frame choose all the rows where the CR_exp_moratoria variable is 0.
   (j) From the resulting data frame, drop all rows where the Amount variable is 0.


3. Select the leading digits of the amounts and compare them with the theoretical frequencies
from Benford's law. Do the data look ok or suspicious?

```
1. Go to the EBA website https://www.eba.europa.eu/risk-analysis-and-data/eu-wide-stress-testing
and download the 
file Credit Risk IRB (https://www.eba.europa.eu/assets/st21/full_database/TRA_CRE_IRB.csv).
You can do so by downloading the file, storing it locally and then read it into R or you can
read it directly from the web.

```

We download the file directly and store it in an R-object, which we call `eba_data`

```{r download-eba}
eba_data <- read.csv("https://www.eba.europa.eu/assets/st21/full_database/TRA_CRE_IRB.csv", stringsAsFactors = FALSE)
```

This should result in an R-object `eba_data` (or whatever name you chose to give to the data) with
528550 records of 15 variables. One way to check this would be the `dim()` function we also
used in the lecture:

```{r check-data-dim}
dim(eba_data)
```

```
2-(a) Extract all variables names, using the `names()`function.

```
We refer to our data object by name `eba_data`. You must choose the name you have selected, of course.

```{r extract-names}
names(eba_data)
```

```

2-(b) Select all rows where the Scenario variable has value 1. Note
       that the symbol you need in the R syntax for equal is `==`, the syntax is therefore 
       `Scenario == 1`. You might check out the R-help entry `Comparison` for further details.
       
```
We overwrite our object name `eba_data`at each step.

```{r scenario filter}
eba_data <- eba_data[eba_data$Scenario == 1, ]
```

Explanation: Since our object name is `eba_data` we subset it by `eba_data[, ]$. The
first slot adresses rows the scond slot addresses columns. We want to say: Take those rows
where the column Scenario tales value 1. The column scenario is `eba_data$Scenario`. The
filter condition is thus written in the row slot. The column slot is left blank because we
want to have records for each variable. Note that without warning R has created here a new
object, with the same name than the old one. But the new `eba_data` has only those rows left,
where the Scenario-Variable is euqal 1.

```
2 -(c) From the resulting data-frame select all rows where the Country variable is not equal to 0.        (hint: The not equal operator in the R syntax is `!=`). If you look into the Metadata-File
       you will see that 0 are all the aggregate exposures not broken down by country. Excluding 
       these will give us country exposures.
```

```{r country-select}
eba_data <- eba_data[eba_data$Country != 0, ]
```

```
2 -(d) From the resulting data frame select all rows where the Portfolio variable has value 1 or         2.These codes describe the accounting rules under which the exposure values are reported, 
       internal rating based (IRB) or standard approach (SA). As a hint you can use R's subset
       operator `%in%` here so `Portfolio %in% c(1,2)` written with the approprate subsetting rule
       will select all rows where the Porfolio variable has value 1 or 2.
```

```{r portfolio-select}
eba_data <- eba_data[eba_data$Portfolio %in% c(1,2), ]
```

```
2-(e) From the resulting data frame choose all the rows where the Exposure variable is not 0.This
       gives again disaggregated numbers.
```

```{r exposure-select}
eba_data <- eba_data[eba_data$Exposure != 0, ]
```

```
2-(f) From the resulting data frame choose all the rows where the Status variable has value 0.

```

```{r filter-status}
eba_data <- eba_data[eba_data$Status == 0, ]
```

```
2-(g) From the resulting data frame choose all the rows where the IFRS9_Stages variable has value        0.
```
```{r filter IFRS}
eba_data <- eba_data[eba_data$IFRS9_Stages %in% c(1,2,3), ]
```

```
2-(h) From the resulting data frame choose all the rows where the CR_guarantees variable is 0

```
```{r filter CR_guarantees}
eba_data <- eba_data[eba_data$CR_guarantees == 0, ]
```

```
2- (i) From the resulting data frame choose all the rows where the CR_exp_moratoria variable is 0.

```

```{r filter-smth}
eba_data <- eba_data[eba_data$CR_exp_moratoria == 0, ]
```

```
 2- (j) From the resulting data frame, drop all rows where the Amount variable is 0.
   
```

```{r filter-zero-am}
eba_data <- eba_data[eba_data$Amount!= 0, ]
```

This results in a new data frame with dimension:

```{r dim-new}
dim(eba_data)
```

