---
title: 'Project 1: Craps, a casino dice game'
author: "Martin Summer"
date: "5.1.2022"
output:
  html_document:
    df_print: paged
---

# Project description

The first project is about one of the very few dice games offered at a Casino: 
Craps.^[ This will be
the only project in this course where we do not go into a direct financial application. 
Before we can do so, we need to build still a bit more knowledge of probability and
skills in R.]

In Craps people throw a pair of dice and bet on outcomes of these
throws. In this project we practice our new knowledge by
implementing the dice throws of craps and the automatic summation of points on the computer.
We try to find out the 
probability of the possible outcomes and check the fairness of the dice.

1. Your first task in this project will be to write a function which allows you 
to virtually throw a pair of dice and sum up the points shown after the throw.
2. Simulate and plot the result of your simulation using `qplot()` with the `binwidth`
argument set to 1.
3. Are these dice fair? Why?
4. Assume the dice were unfair in the following sense: Numbers 1,2,3,4 and 5 have a 
probability of $1/8$ while the 6 has a probability of $3/8$. Study the help page of
the `sample`function and find out how you could give these new probability weights
to the function. If you redo your simulation analysis with the loaded dice, how does the
picture change?
5. Write a script for this random experiment.
6. Now look at the random experiment of throwing two dice with the concepts from
probability theory: What is the sample space of this experiment. What are the 
probabilities of the basic outcomes? What is the probability of getting sum 7, what is
the probability of getting sum 2?
7. In the early 17th century the Grand Duke of Tuscany was puzzled because he thought that
counting possible cases in a game where three dice are thrown seemed 
to give the wrong answer. Counting combinations of numbers, 10 and 11 can be 
made in 6 ways, as can 9 and 12", yet experienced dice players regard 10 and 11 
to be more advantageous than 9 and 12. "How 
can this be ?", the count asked Galileo Galilei (1564 - 1642) 
who composed a short note to explain to the
count what was going on. What might have been Galileo's reply?
8. Imagine now that you are at the casino and you can spend 150 Euro
for bets on single rolls of  the two dice. 
Since you have figured out that 7 is the most
likely outcome you would like to buy bets on 7. The casino
offers you a bet for 0.15 cent, so you could make 1000 bets 
on 7. Is this a good or a bad deal for you? Try to think about
this problem in terms of a simulation.

```
1. Your first task in this project will be to write a function which allows you 
to virtually throw a pair of dice and sum up the points shown after the throw. You should write this function as a script. A script can be created in R-Studio by clicking
`File > new File > R Script`. Here is a suggestion how the code could look like. Note that
a particular functionality can be implemented in several ways in R. So the "solution"
presented here should be regarded as a suggestion only.

```
Here is one possible implementation. The lines starting with
`#` are comments. You can read them when editing the function script
but R will igore them when interpreting the function.

```{r function to roll two dice and sum their points}
roll <- function(){
  
  # create a die
  
  die <- 1:6
  
  # roll the dice by making use of sample. We roll two dice, therefore size has to be equal to 2
  # note that we need to set replace to TRUE because both dice should be able to show all possible
  # points
  
  dice <- sample(die, size = 2, replace = TRUE)
  
  # sum up the points of the two dice
  
  sum(dice)
  
}
```

```
2. Simulate and plot the result of your simulation using `qplot()` with the `binwidth`
argument set to 1.

```

```{r simulate and plot}

rolls <- replicate(10000, roll())

# load library ggplot2. Note: you can load ggplot2 only if you have installed the package first:
# This can be done by typing install.packages(ggplot2) at the prompt of the R console or
# use the RStudio menues in the lower right panel.

library(ggplot2)

qplot(rolls, binwidth = 1)

```

```
3. Are these dice fair? Why?

```

We answer this question by considering the sample space of this random experiment first.
There are in total $6 \times 6$ pairs, in total 36 ways to sum up points. Each pair
has probability $1/36$ and since the multiple pairs do not occur in the same
event the pair probabilities that lead to a particular sum can be added. The sample space
${\cal S}$ can be visualized like this:

| Probability         | Outcome              | Possibilities                      |
|:-------------------:|:--------------------:|:-----------------------------------|
|P(sum =  2)  = 1/36  | 2                    | (1,1)                              |
|P(sum =  3)  = 2/36  | 3                    | (1,2),(2,1)                        |
|P(sum =  4)  = 3/36  | 4                    | (1,3),(2,2),(3,1)                  |
|P(sum =  5)  = 4/36  | 5                    | (1,4),(2,3),(3,2),(4,1)            |
|P(sum =  6)  = 5/36  | 6                    | (1,5),(2,4),(3,3),(4,2),(5,1)      |
|P(sum =  7)  = 6/36  | 7                    | (1,6),(2,5),(3,4),(4,3),(5,2),(6,1)|               |P(sum =  8)  = 5/36  | 8                    | (2,6),(3,5),(4,4),(5,3),(6,2)      |
|P(sum =  9)  = 4/36  | 9                    | (3,6),(4,5),(5,4),(6,3)            |
|P(sum = 10)  = 3/36  |10                    | (4,6),(5,5),(6,4)                  |
|P(sum = 11)  = 2/36  |11                    | (5,6),(6,5)                        |
|P(sum = 12)  = 1/36  |12                    | (6,6)                              | 

The frequencies with wich the sums occur look very much like the theoretical probabilities
for equally probable pairs. Our dice thus look as if they are fair.

```
4. Assume the dice were unfair in the following sense: Numbers 1,2,3,4 and 5 have a 
probability of $1/8$ while the 6 has a probability of $3/8$. Study the help page of
the `sample`function and find out how you could give these new probability weights
to the function. If you redo your simulation analysis with the loaded dice, how does the
picture change?

```
```{r biased-dice}
roll_biased <- function(){
  
  # Construct die
  die <- 1:6
  
  # Roll biased or laded die by specifying the probabilities such that there is relatively
  # more mass on outcome 6.
  dice <- sample(die, size = 2, replace = TRUE, prob = c(1/8, 1/8, 1/8, 1/8, 1/8, 3/8))
  
  sum(dice)
}
```

```{r simulate-biased-die}
rolls <- replicate(10000, roll_biased())

qplot(rolls, binwidth = 1)
```
```
5. Write a script for this random experiment.

```

We create a script with `File > new File > R script`and write the code for `roll_biased` into the script. That's it. 

```
6. Now look at the random experiment of throwing two dice with the concepts from
probability theory: What is the sample space of this experiment. What are the 
probabilities of the basic outcomes? What is the probability of getting sum 7, what is
the probability of getting sum 2?

```

The sample space is:
${\cal S} = \{ (1,1), (1,2), (1,3), (1,4), (1,5), (1,6), (2,6), (3,6), (4,6), (5,6), (6,6),
(2,1), (2,2), (2,3), (2,4), (2,5), (3,5), (4,5), (5,5), (6,5), (3,1), (3,2), (3,3), (3,4), 
(4,4), (5,4), (6,4), (4,1), (4,2), (4,3), (5,3), (6,3), (5,1), (5,2), (6,2), (6,1) \}$

The probabilities of the basic outcomes are $1/36$. 

The probability of getting the sum 7 is
$6/36$ or $1/9$, 
the probability of getting a 2 is $1/36$.

```

7. In the early 17th century the Grand Duke of Tuscany was puzzled because he thought that
counting possible cases in a game where three dice are thrown seemed 
to give the wrong answer. Counting combinations of numbers, 10 and 11 can be 
made in 6 ways, as can 9 and 12", yet experienced dice players regard 10 and 11 
to be more advantageous than 9 and 12. "How 
can this be ?", the count asked Galileo Galilei (1564 - 1642) 
who composed a short note to explain to the
count what was going on. What might have been Galileo's reply?

```

Great count of Tuscany, you are counting the wrong thing. You count $(3,3,3)$ as one 
possibility of making a $9$ and two $3$ and one $4$ as one possibility of making a $10$. But the
latter covers three possibilities $(4,3,3), (3,4,3), (3,3,4)$. 
For the former it is only $(3,3,3)$. You need to take into account permutations and
combinations ans the puzzle disappears.

```
8. Imagine now that you are at the casino and you can spend 200 Euro
for chips. Since you have figured out that 7 is the most
likely outcome you would like to buy bets on 7. The casino
offers you a bet for 20 cent. This means you pay 20 cents
for each roll of the die and when 7 comes up you get 1 Euro, when
another sum comes up you get nothing. With your budget
you could make 1000 bets 
on 7. Is this offer of the casion 
a good or a bad deal for you? Try to think about
this problem in terms of a simulation. 

```
With 15 cent per bet and a budget of 150 we can bet on 1000 rolls
of the dice. 

```{r simulate a craps bet}

# simulate 1000 oucomes with fair dice

rolls <- replicate(1000, roll())

# count the rolls where the outcome is 7.For these rolls you get 1.
# Since you pay 0.2 at each roll you spend your budget of 200
# over the course of the rolls. Your profit or loss computes as

sum(rolls == 7) - 200

```
No, you would end the game with a loss and pay 33 in addition to the
200 out of your pocket to the Casino.

If you made the same bet with a bet price of 10 cents you would get over a course of 1000 rounds.

```{r cheaper ticket}
sum(rolls == 7) - 100
```


