
---
title: "Exercises and Project: Lecture 1 Solutions"
format: html
---

## Exercises with Solutions

### Exercise 1: Simulate Rolling Two Dice

#### Task
Write a function to simulate rolling a pair of dice. Simulate 10,000 rolls and visualize the distribution of sums.

#### Solution
```{r}
# Function to roll two dice and sum their outcomes
roll_dice <- function() {
  sum(sample(1:6, size = 2, replace = TRUE))
}

# Simulate 10,000 rolls
set.seed(123)
dice_rolls <- replicate(10000, roll_dice())

# Visualize the results
library(ggplot2)
dice_data <- data.frame(Sum = dice_rolls)
ggplot(data = dice_data, aes(x = Sum)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Sums of Two Dice",
    x = "Sum",
    y = "Frequency"
  ) +
  theme_minimal()
```

---

### Exercise 2: Analyze the Fairness of Dice

#### Task
Compute the theoretical probabilities of sums for two fair dice and compare with simulated frequencies.

#### Solution
```{r}
# Theoretical probabilities
fair_probs <- data.frame(
  Sum = 2:12,
  Probability = c(1, 2, 3, 4, 5, 6, 5, 4, 3, 2, 1) / 36
)

# Simulated probabilities
simulated_probs <- as.data.frame(table(dice_rolls) / length(dice_rolls))
colnames(simulated_probs) <- c("Sum", "Simulated_Probability")
simulated_probs$Sum <- as.numeric(as.character(simulated_probs$Sum))

# Merge and compare
comparison <- merge(fair_probs, simulated_probs, by = "Sum")
comparison
```

**Comments**:
- Simulated probabilities should closely match theoretical probabilities if the dice are fair.

---

### Exercise 3: Simulate Loaded Dice

#### Task
Simulate rolls for loaded dice where 1-5 have a probability of 1/8, and 6 has a probability of 3/8. Visualize the results.

#### Solution
```{r}
# Function to roll loaded dice
roll_loaded_dice <- function() {
  prob <- c(1/8, 1/8, 1/8, 1/8, 1/8, 3/8)
  sum(sample(1:6, size = 2, replace = TRUE, prob = prob))
}

# Simulate 10,000 rolls
set.seed(123)
loaded_rolls <- replicate(10000, roll_loaded_dice())

# Visualize
loaded_data <- data.frame(Sum = loaded_rolls)
ggplot(data = loaded_data, aes(x = Sum)) +
  geom_histogram(binwidth = 1, fill = "pink", color = "black") +
  labs(
    title = "Distribution of Sums of Loaded Dice",
    x = "Sum",
    y = "Frequency"
  ) +
  theme_minimal()
```

**Comments**:
- Observe how the distribution shifts due to the loading. Sum 12 becomes more likely due to the increased probability of rolling 6.

---

## Project with Solutions

### Project: Casino Dice Game and Probabilities

#### Exercise 4: Probability of Specific Sums

**Task**: Calculate the probability of rolling a sum of 7 and 2 for both fair and loaded dice.

**Solution**:
```{r}
# Fair dice probabilities
prob_sum_7_fair <- 6 / 36
prob_sum_2_fair <- 1 / 36

# Loaded dice probabilities
prob_sum_7_loaded <- (5 * 3 + 5 * 3) / 64
prob_sum_2_loaded <- (1 * 1) / 64

list(
  "Fair Dice" = list("Sum 7" = prob_sum_7_fair, "Sum 2" = prob_sum_2_fair),
  "Loaded Dice" = list("Sum 7" = prob_sum_7_loaded, "Sum 2" = prob_sum_2_loaded)
)
```

**Comments**:
- Sum 7 remains most likely for fair dice but changes significantly for loaded dice.

---

#### Exercise 5: Evaluating a Betting Strategy

**Task**: Simulate 1,000 bets on sum 7 for both fair and loaded dice.

**Solution**:
```{r}
simulate_bets <- function(num_bets, cost_per_bet, payout, prob_win) {
  results <- sample(c(payout, -cost_per_bet), num_bets, replace = TRUE, prob = c(prob_win, 1 - prob_win))
  sum(results)
}

# Fair dice
fair_profit <- replicate(10000, simulate_bets(1000, 0.15, 1, prob_sum_7_fair))
mean(fair_profit)

# Loaded dice
loaded_profit <- replicate(10000, simulate_bets(1000, 0.15, 1, prob_sum_7_loaded))
mean(loaded_profit)
```

**Comments**:
- Discuss expected values and profitability under fair and loaded conditions.
- Highlight how casinos use odds to maintain their advantage.

---

### Conclusion
This set of exercises and project demonstrates how probability principles and R simulations can be applied to understand real-world scenarios like dice games and casino strategies. Through this analysis, students gain hands-on experience in probability modeling and critical thinking about fairness and profitability.
