---
title: "Evaluating Credit Risk Using Conditional Probabilities"
format: html
execute:
  echo: true
---

## Project: Evaluating Credit Risk Using Conditional Probabilities

### Problem Description

A bank is evaluating a loan application using historical data to estimate the likelihood of default. The borrower has a low credit score, and the bank has the following data:

1. **Default Rates**:
   - Probability of default for all customers: $P(D) = 0.04$.
   - Probability of non-default for all customers: $P(ND) = 0.96$.

2. **Evidence**:
   - Probability of a low credit score given default: $P(L|D) = 0.7$.
   - Probability of a low credit score given non-default: $P(L|ND) = 0.1$.

The goal is to determine the posterior probability of default given the borrower’s low credit score, $P(D|L)$, using **Bayes' Rule**. Additionally, we will verify this theoretical result by simulating customer data and analyzing outcomes.

---

### Questions

1. **Compute $P(D|L)$ Theoretically**:
   - Use Bayes' Rule to calculate the posterior probability of default given a low credit score.

2. **Simulate the Scenario in R**:
   - Simulate a dataset of 10,000 customers where each customer is randomly assigned a default status based on $P(D)$.
   - Based on the assigned default status, simulate whether each customer has a low credit score using $P(L|D)$ and $P(L|ND)$.

3. **Compute $P(D|L)$ from Simulated Data**:
   - Use the simulated data to compute $P(D|L)$ and compare it to the theoretical result.

4. **Visualize Results**:
   - Create a simple bar plot comparing the simulated and theoretical probabilities.

---

## Solution

### Question 1: Compute $P(D|L)$ Theoretically

Using Bayes' Rule, we can compute $P(D|L)$ as:

$$P(D|L) = \frac{P(L|D) \cdot P(D)}{P(L)}$$

where:

$$P(L) = P(L|D) \cdot P(D) + P(L|ND) \cdot P(ND)$$

Substitute the given probabilities:

```{r}
# Given probabilities
P_D <- 0.04              # P(D)
P_ND <- 0.96             # P(ND)
P_L_given_D <- 0.7       # P(L|D)
P_L_given_ND <- 0.1      # P(L|ND)

# Compute P(L)
P_L <- P_L_given_D * P_D + P_L_given_ND * P_ND

# Compute P(D|L)
P_D_given_L <- (P_L_given_D * P_D) / P_L

# Output the result
cat("Theoretical P(D|L):", P_D_given_L, "\n")
```

### Question 2: Simulate the Scenario in R

We simulate a dataset of 10,000 customers without using binomial distributions or conditional statements:

```{r}
# Set seed for reproducibility
set.seed(123)

# Number of customers
n_customers <- 10000

# Generate random numbers for default assignment
random_defaults <- runif(n_customers)
default_status <- numeric(n_customers)

# Assign defaults based on P(D)
for (i in 1:n_customers) {
  if (random_defaults[i] <= P_D) {
    default_status[i] <- 1  # Default
  } else {
    default_status[i] <- 0  # Non-Default
  }
}

# Generate random numbers for low credit score assignment
random_low_credit <- runif(n_customers)
low_credit <- numeric(n_customers)

# Assign low credit scores based on default status
for (i in 1:n_customers) {
  if (default_status[i] == 1) {
    if (random_low_credit[i] <= P_L_given_D) {
      low_credit[i] <- 1  # Low Credit Score
    } else {
      low_credit[i] <- 0
    }
  } else {
    if (random_low_credit[i] <= P_L_given_ND) {
      low_credit[i] <- 1  # Low Credit Score
    } else {
      low_credit[i] <- 0
    }
  }
}

# Create a data frame of simulated customers
simulated_data <- data.frame(
  Default = default_status,
  LowCredit = low_credit
)

# Display the first few rows
head(simulated_data)
```

### Question 3: Compute $P(D|L)$ from Simulated Data

Using the simulated data, calculate the empirical $P(D|L)$:

```{r}
# Filter customers with low credit scores
low_credit_customers <- simulated_data[simulated_data$LowCredit == 1, ]

# Compute the proportion of low-credit customers who defaulted
empirical_P_D_given_L <- mean(low_credit_customers$Default)

# Output the result
cat("Empirical P(D|L):", empirical_P_D_given_L, "\n")
```

### Question 4: Visualize Results

We compare the theoretical and empirical probabilities using a bar plot:

```{r}
# Create a vector of probabilities
probabilities <- c(Theoretical = P_D_given_L, Empirical = empirical_P_D_given_L)

# Bar plot
barplot(probabilities, 
        col = c("blue", "green"), 
        ylim = c(0, 1), 
        main = "Comparison of Theoretical and Empirical P(D|L)",
        ylab = "Probability",
        names.arg = c("Theoretical", "Empirical"))

# Add legend
legend("topright", 
       legend = c("Theoretical", "Empirical"), 
       fill = c("blue", "green"))
```

---

## Discussion

### Key Observations:

1. **Theoretical vs. Empirical Results**:
   - Theoretical $P(D|L)$ was derived using Bayes' Rule and represents the expected result if all probabilities are accurate.
   - Empirical $P(D|L)$ closely aligns with the theoretical result, validating the simulation process.

2. **Significance of Conditional Probability**:
   - This project demonstrates how conditional probabilities can provide valuable insights for credit risk evaluation.
   - A low credit score significantly increases the likelihood of default, as indicated by the posterior probability $P(D|L)$.

3. **Practical Applications**:
   - Banks can use similar methods to tailor lending decisions, focusing on customers with high $P(D|L)$.

4. **Limitations**:
   - The simulation assumes probabilities are fixed and accurately represent the population.
   - Real-world scenarios may include additional factors (e.g., economic conditions, borrower history) that influence default rates.

---
