---
title: "My coin tossing function"
author: "Martin Summer"
editor: source
---

## Tossing a Coin in R {#coin-tossing}

   The function `toss_coin()` allows us to simulate the random experiment of tossing a coin. 
   Here is the code:
   
```{r}
toss_coin <- function() {
       coin <- 0:1
       sample(coin, size = 1, replace = TRUE)
   }
```
   
### Explanation of the Code

   The `toss_coin` function uses the `sample` function to randomly select one of the two values in the vector `coin`. Setting `replace = TRUE` ensures that each toss is independent. Here's an example of a series of tosses:

```{r}
tosses <- replicate(10, toss_coin())
   tosses
```

The output is a vector of 10 random toss outcomes.

```{r}
# Load the ggplot2 package
library(ggplot2)

# Simulate 100 coin tosses
tosses <- replicate(100, sample(c(0, 1), size = 1, replace = TRUE))

# Create a data frame for plotting
tosses_df <- data.frame(
  Outcome = factor(tosses, levels = c(0, 1), labels = c("Tails", "Heads"))
)

# Plot the results using ggplot2
ggplot(tosses_df, aes(x = Outcome)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Occurrences of Heads and Tails",
    x = "Outcome",
    y = "Count"
  ) +
  theme_minimal()

```

```{r}
set.seed(123)
tosses2 <- replicate(10000, toss_coin())
```

```{r}

```

