---
title: "EBA data and Bendofrd's law"
output: html_notebook
author: Martin Summer
date: 9.1.2021
---

```{r}
eba <- read.csv("https://www.eba.europa.eu/assets/st21/full_database/TRA_CRE_IRB.csv")
View(eba)
```

Here is what you are asked to do: We look only at the actual exposure data, which is
coded as all rows where Scenario is equal to 1. We look at the data broken down by Country, which
means all rows where Country is not equal to 0. We want no portfolio break downs and look only
at rows where Portfolio is equal to 0, we don't look at an exposure break down and thus set
the variable Exposure to 0. We also set Status, IFRS9_Status, CR_guarantees and CR_exp_moratoria should also be set to zero. Construct a new dataframe, with these filtered data.

```{r}
eba_filtered <- eba[(eba$Scenario == 1 &
                     eba$Country !=0 &
                     eba$Portfolio %in% c(1,2) &
                     eba$Exposure == 0 &
                     eba$Status == 0 &
                     eba$IFRS9_Stages %in% c(1,2,3) &
                     eba$CR_guarantees == 0 &
                     eba$CR_exp_moratoria == 0 &
                     eba$Amount != 0), ] 
                
```

```{r}
eba_filtered$Amount <- as.numeric(eba_filtered$Amount)*10^6

eba_filtered <- na.omit(eba_filtered)

eba_filtered$Amount <- as.character(eba_filtered$Amount)
```

```{r}
eba_filtered$LD <- substr(eba_filtered$Amount,1,1)
```

```{r}
tab <- as.data.frame(table(eba_filtered$LD)/length(eba_filtered$LD))
tab$Freq_Benf <- log10(1 + 1/as.numeric(tab$Var1))
tab
```

