\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Lecture 5: Monte Carlo Methods and Applications}
\author{}
\date{}

\begin{document}
	
	\maketitle
	
	\section*{Lecture 5 Outline: Monte Carlo Methods and Applications}
	
	\section*{1. Introduction to Continuous Random Variables}
	\begin{itemize}
		\item Brief recap of key concepts: continuous vs. discrete random variables, the normal distribution.
	\end{itemize}
	
	\section*{2. Lognormal Distribution}
	\begin{itemize}
		\item Introduce the lognormal distribution, emphasizing its application in modeling stock prices.
		\item \textbf{Practical Exercise in R:} Simulate and visualize lognormal distributions and stock prices.
	\end{itemize}
	
	\section*{3. Monte Carlo Methods and Applications}
	\begin{itemize}
		\item \textbf{Monte Carlo Simulation Basics:}
		\begin{itemize}
			\item Introduce the concept of Monte Carlo simulations and their use cases in finance.
			\item \textbf{R Example:} Simulate portfolio returns and estimate risk metrics (e.g., Value at Risk).
		\end{itemize}
		\item \textbf{Optimizing Monte Carlo Simulations in R:}
		\begin{itemize}
			\item Introduce key R tools and techniques for speeding up code:
			\begin{itemize}
				\item Vectorization.
				\item Parallel processing (e.g., using \texttt{future} package).
				\item Efficient data handling with \texttt{data.table}.
			\end{itemize}
			\item \textbf{Hands-on Exercise:} Compare a simple brute-force Monte Carlo simulation with an optimized version.
		\end{itemize}
	\end{itemize}
	
	\section*{4. Risk Management Applications}
	\begin{itemize}
		\item \textbf{Value at Risk (VaR):}
		\begin{itemize}
			\item Expand on VaR calculations and their assumptions.
			\item Discuss handling deviations from normality (e.g., fat tails, alternative distributions).
			\item \textbf{Case Study:} Use Monte Carlo simulations to estimate VaR for a stock portfolio.
		\end{itemize}
	\end{itemize}
	
	\section*{5. Estimation Issues and Real-World Data}
	\begin{itemize}
		\item \textbf{Statistical Estimation Challenges:}
		\begin{itemize}
			\item Discuss estimating expected returns, variance, and covariance from data.
			\item Address practical issues like statistical noise.
			\item \textbf{R Exercise:} Estimation techniques with real-world stock return data.
		\end{itemize}
	\end{itemize}
	
	\section*{6. R Programming: Efficiency and Documentation}
	\begin{itemize}
		\item \textbf{Optimizing Code in Practice:}
		\begin{itemize}
			\item Introduce best practices for writing efficient and maintainable R code.
			\item Tools for documenting code (\texttt{Roxygen2}), version control, and team collaboration.
			\item \textbf{Exercise:} Develop a small, well-documented R function related to Monte Carlo simulations.
		\end{itemize}
	\end{itemize}
	
	\section*{7. References and Further Study}
	\begin{itemize}
		\item Allocate time at the end for additional resources, textbooks, and suggestions for further study in financial modeling and risk management.
	\end{itemize}
	
\end{document}
